---
title: "Data manipulation"
output:
  html_document:
    toc: true 
    toc_float: true
    toc_collapsed: true
    toc_depth: 3  
layout: default
---

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

### Group_by and summarize

So far, we have only been summarizing on specific columns, but often we are interested in summaries based on different groups. We can do this utilizing both the `group_by()` and `summarize()` functions. `group_by()` puts the data frame in "groups" based on a column or columns. This is down "behind the scenes" and doesn't change the data output. But, when your run other functions, like `mutate()` or `summarize()` the groups are honored. 

`summarize()` allows you to make new columns based on a summary functions -- like the mean, median, or count. `summarize()` uses a similar format to `mutate`, new_column = function. 

For example, if I wanted to get the average (`mean()`) of mean temperature (tmean) for the different months, you could do the following:

```{r}
climate |>
  group_by(year) |>
  summarize(average_temp = mean(tmean)) |>
  head()
```

You can also make multiple summaries at once by adding into the `summarize()` function. Here, we can add in the standard deviation (`sd()`) and average and standard deviation for precipitation. Let's also add in a filter for only years in the 90's.

```{r}
climate |>
  filter(year %in% 1990:1999) |>
  group_by(year) |>
  summarize(average_temp = mean(tmean), sd_temp = sd(tmean), average_precip = mean(precip), sd_precip = sd(precip))
```

[**Exercise**](#summarizing-data): Which year had the highest *total* precipitation in Raleigh?


## Water quality data

In this course, we will exploring historic water quality data from the City of Raleigh. The data is collected from 2008 - 2023 and includes 19 different water quality variables, including water temperature, *E. coli* bacteria, and dissolved oxygen for 18 different stream sites. 

Let's now go back to our City of Raleigh water quality data and use plots to explore the data. 

First, we need to read in the data into R again. 

```{r}
wq_clean <- read.csv('https://maryglover.github.io/bio331/water_quality_manip/raleigh_wq_clean.csv')
```

**Activity: **First, pick one of the different parameters and make a histogram. Record the code in your script and make a note about what you see. 

Now, let's use the data to explore the relationship of temperature and dissolved oxygen (in percent saturation). What do you predict?

```{r}
ggplot(wq_clean, aes(x = Temperature_C, do_percent_sat)) +
  geom_point()
```  

Now, let's see what we notice about the different sites. We can start with exploring the amount of phosphorus at each site using a boxplot.

```{r}
ggplot(data = wq_clean, aes(x = Site, y = Phosphorus_total_mg_L )) +
  geom_boxplot()
```

You can see that there are duplicate sites! This is part of the City of Raleigh's quality control. In some cases, they will collect duplicates as a way to make sure they are getting accurate results. 

Run the following code to remove the duplicates. You don't need to understand what this code is doing for assessments, but it will filter out any site with the "DUP" or "Dup" in them. 

```{r}
wq_clean<- wq_clean |>
  filter(!grepl("DUP",Site)) |> # gets rid of duplicate sites
  filter(!grepl("Dup",Site))
```

Now, rerun the boxplot code. 

```{r}
ggplot(data = wq_clean, aes(x = Site, y = Phosphorus_total_mg_L )) +
  geom_boxplot()
```

Now, look at the amount of E. coli in MPN units. What do you notice?

```{r}
ggplot(data = wq_clean, aes(x = Site, y = E_coli_MPN_100mL )) +
  geom_boxplot()
```

Here, you can see that there is an outlier! Use `dplyr` functions to sort the *E. coli* data to look more closely. 

This is an example of a time where good scientific judgement is needed to figure out what to do with this data point. It is *not* appropriate to remove a data point just because it is very different from another. However, there are times when an outlier is due to an error in data collection or recording. It is important to understand the parameter to determine what is going on. This is more complicated when we did not collect the data. In this instance, I was able to talk to the data collector and they indicated that this value for *E. coli * is not a reasonable value and is likely due to a lab issue. In this case, we can remove this data point. 

Run the following code to do that. 

```{r}
wq_clean <- wq_clean |>
 mutate(E_coli_MPN_100mL = gsub('155310', NA, E_coli_MPN_100mL))
```

Because we have made some changes in the water quality data, it is a good idea to save the data set. We can do this with the `write.csv()` function. We *should not* rewrite the orignial data file, but should rename it. Here, we provide the arguments for the data frame we want to save, and the name of the file. 

```{r, eval=FALSE}
write.csv(wq_clean, 'data/wq_clean2.csv', row_names = FALSE)
```
