---
title: "Summary Tables"
output:
  html_document:
    toc: true 
    toc_float: true
    toc_collapsed: true
    toc_depth: 3  
layout: default
---

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

You can use this [template script](https://maryglover.github.io/bio331/lessons/summaries/summary_script.R) file to work through the lesson. 

## Creating summary tables

We have begun in class summarizing different variables in data frames. But often we are interested in summaries based on different groups. For example, in our climate data analysis, we looked to see what the average precipitation value was in Raleigh, but maybe we want to see the average for every year. We can do this utilizing both the `group_by()` and `summarize()` functions. 

- `group_by()` puts the data frame in "groups" based on a column or columns. This is done "behind the scenes" and doesn't change the data output. But, when your run other functions, like `mutate()` or `summarize()` the groups are honored. 

- `summarize()` allows you to make new columns based on a summary functions -- like the mean, median, or count. `summarize()` uses a similar format to `mutate`, new_column = function. 

First, let's load in our climate data and our dplyr package. 

```{r, message = F}
climate <- read.csv('data/raleigh_prism_climate.csv')
library(dplyr)
```

Now, I will **summarize** to get the *mean* of precipitation  (`precip`), **grouping** for the different years:

```{r, message = F}
climate |>
  group_by(year) |>
  summarize(average_precip = mean(precip)) 
```

This first code `group_by` groups the data by the year, so that when you do the `summarize` the average will be taken for each year group. 

You can also make multiple summaries at once by adding more calculations into the `summarize()` function. Here, we can add in the standard deviation (`sd()`). You can add as many summary calculations as you would like. Let's also add in a filter for only years in the 90's to simply the table.

```{r}
climate |>
  filter(year %in% 1990:1999) |>
  group_by(year) |>
  summarize(average_precip = mean(precip), sd_precip = sd(precip))
```

<span style="color:blue">**Activity:**</span>

- What year had the highest average temperature `tmean`?
- Which year had the highest *total* precipitation in Raleigh?

## More examples

### Raleigh cycling crashes

The City of Raleigh keeps track of incidents that involve bicycles and vehicles in Raleigh. We will use this data set to practice making summaries. 

First, let's load in the data, which I have stored in our class website. Here, I am assigning it the name `cycling`.

```{r}
cycling <- read.csv('https://maryglover.github.io/bio331/open_data/cycling_crashes.csv')
```

Let's look at the data, by viewing the first few rows and the column names. 

```{r}
head(cycling)
colnames(cycling)
```

We can see that this data set includes the date of the incident (day_of_week, day, month, and year), where it occurred (road), and different characteristics (road_feature, weather, traffic control). 

Let's say we want to see how many crashes occurred each year. To get the number of crashes, we can want to know the number of rows in each group. This data is in the "tidy" format, so each crash is one row. In `dplyr` we can simplify this  by using the function `n()`, which gives the number of items in the specified group. 

```{r}
cycling |>
  group_by(year)|>
  summarise(number = n())
```

Now, let's say we wanted to look at the number of crashes based on the weather, to see if the weather causes more crashes. Instead of grouping by the year, we simply change the group to weather.

```{r}
cycling |>
  group_by(weather)|>
  summarise(number = n())
```

<span style="color:blue">**Activity:**</span>

- Summarize the data to determine the number of crashes for each road and *then* sort to see what road had the most crashes. 

```{r, include = F, echo = F}
cycling |>
  group_by(road) |>
  summarize(n = n())|>
  arrange(-n) |>
  head()
```

### Names data from birth records

In our next example, we are using a dataset with names of babies born from years 1910 to 2024. Use the following code to read in the dataset and assigning it the name `birth`

```{r}
birth <- read.csv('https://maryglover.github.io/bio331/open_data/nc_birth.csv')
```

Look just at the first few rows 
```{r}
head(birth)
```

You can see that it has the number of babies born in a year with different names. It also includes a column indicating the decade. 

Let's first summarize to what is the most common name for boys and girls for the entire data set. In this instance, we will group by **both**  the name and the sex. You can do this by adding both of the columns in to the `group_by` function. Because there are so many names, let's also arrange the data to sort by the most common. 

```{r, warning = F, message = F}
birth |>
  group_by(sex, name) |>
  summarize(total = sum(number)) |>
  arrange(-total)
```

We can also use the decade column to summarize data. Let's say we want to know the most common name for girls for each decade. First, we will filter to just get data for girls. Then, we can group for both decade *and* name. Lastly, we will summarize for the total babies born with each name. 

```{r, message=F}
birth |>
  filter(sex == 'F')|>
  group_by(decade, name) |>
  summarize(total = sum(number))
```

We now have the number of times a name was given in every decade. We can use the useful function `slice_max` which we give us just the row that has the highest value in a specific column. We only want the top max value, so can set `n=1` in `slice_max()`. We want where the `total` value is the highest. 

```{r}
birth |>
  filter(sex == 'F')|>
  group_by(decade, name) |>
  summarize(total = sum(number)) |> 
  slice_max(total, n = 1)
```  

<span style="color:blue">**Activity:**</span>

- Filter the data for the 2000's decade and determine the top 5 names for boys and girls. 

```{r, include = F, echo = F}
birth |>
  filter(decade == '2000') |>
  group_by(sex, name) |>
  summarize(total = sum(number))|>
  slice_max(total, n = 5)
```

### Invertebrate data

For our last example, we will use data collected by this class in 2024 for the macroinvertebrate insects they found during a class period in a nearby stream. 
First, we will read in the data from our class website. I will call it `invert`

```{r}
invert <- read.csv('https://maryglover.github.io/bio331/open_data/macroinvertebrate_sampling.csv')
```

Let's again start by just looking at the data

```{r}
head(invert)
```

In this data, we have the different taxa that two different student groups collected and how many of each they collected. 

Let's say we want to know how many total insects *each* group collected to see if one group collected more than the other. Go ahead and try this summary on your own. You should get the following answer. 

```{r, echo = F}
invert |>
  group_by(Group) |>
  summarize(total = sum(N))
```

Now, let's see how many total insects we collected from each taxa. This gives us the *abundance* of each differnt invertebrate. 

```{r}
invert |>
  group_by(Taxa)|>
  summarize(total = sum(N))
```

One thing you can do with this data is make a bar graph to show the different insects that were collected in the stream. 

First, we want to save the data table we made. Remember, as with other `dplyr` functions, the functions do *not* change the underlying data. 

```{r}
invert_summary <- invert |>
  group_by(Taxa)|>
  summarize(total = sum(N))
```

Then, we can use `ggplot` to plot the data with the layer `geom_bar`.

```{r, message = F}
library(ggplot2)
```

```{r, eval =F}
ggplot(invert_summary, aes(x = Taxa, y = total))+
  geom_bar()
```

You will notice that this code does *not* work. `geom_bar` requires an additional argument that indicates how to determine the bar height. We just want the bar height to be the value in the total column. To do this we must set `stat = "identity"`

```{r}
ggplot(invert_summary, aes(x = Taxa, y = total))+
  geom_bar(stat = 'identity')
```
You will notice that this is hard to read with the axis labels. We can fix this in two ways:

- rotate the axis labels
- plot horizontally 

To rotate the axis labels, you must adjust the angle of the `axis.text.x` in a `theme()` layer like so:

```{r}
ggplot(invert_summary, aes(x = Taxa, y = total))+
  geom_bar(stat = 'identity')+
  theme(axis.text.x = element_text(angle = 90))
```

Alternatively, we can just switch the x and the y to plot in the other direction. 

```{r}
ggplot(invert_summary, aes(y = Taxa, x = total))+
  geom_bar(stat = 'identity')
```
Lastly, let's look at the *species richness* or how many *different* taxa were collected. Here, we want to *tally* the individual taxa in each group. We can do this again with the `n()`
function. 

```{r}
invert |>
  group_by(Group)|>
  summarize(richness = n())
```

## Water quality data

Let's now go back to our City of Raleigh water quality data and use plots to explore the data a bit before we do some summaries. 

First, we need to read in the data into R. Last class, you saved the water quality data, removing the *E. coli* outlier. You will need to read in this data from where you saved it. 

```{r}
wq <- read.csv('data/raleigh_water_analysis.csv')
```

In the water quality data, let's look at how the streams differ in Turbidity. We will calculate the turbidity average for each of the different streams, using the column `Site`.

```{r, eval = F}
wq |>
  group_by(Site) |>
  summarize(mean_turb = mean(Turbidity_NTU))
```

What do you notice?

In this data, and in many other data sets, there are NA values, meaning not applicable or not available. This is used when we do not have information about a value. We can't put zero when we don't know, because we cannot say that the value is zero. Because there are NAs in our data set, we can't say for sure that the maximum turbidity in the streams is not in a stream that wasn't measured. So the answer for the maximum value is NA. To ignore the NAs to get the maximum turbidity based on the values that we measured, we must add `na.rm = TRUE` to the function. 

```{r}
wq |>
  group_by(Site) |>
  summarize(mean_turb = mean(Turbidity_NTU, na.rm = T))
```

Now, what do you notice about the data?

You can use this table to match the Site codes with the stream name. 

```{r, echo = F}
stream_codes <- read.csv('../../data/raleigh_wq_stream_codes.csv', header = F)
knitr::kable(stream_codes)
```

## Homework exercises

Complete the homework exercises in the same R script that you used for class. Submit the R script with both class exercises and homework answers in the Moodle assignment.

### Part 1: Climate data

1. Go back to the climate data. Summarize the data to get the average precipitation for each month. You will group by the month, and summarize to get the mean of the `precip` column. 
2. Repeat to determine the mean and standard deviation (`sd`) of the maximum temperatue (`tmax`) for each month.

### Part 2: Water quality data

1. Go back to the City of Raleigh water quality data. Complete a few more summaries to look at the differences between the sites. *You* choose what summaries you are interested in. Determine what parameters (i.e. do_percent_sat, pH, etc,) and what calculations (i.e. mean, median, max) you want to explore. 
2. Write up a couple of paragraphs of what you notice about the water quality from your summaries. 
3. Write a couple of sentences describing something you would be interested in exploring in the water quality data that you *don't* know who to do.

## List of new functions

- `group_by`
- `summarize`
- `slice_*`
- `sd`
- `geom_bar`
- `na.rm = T`
- `n()`

## Resources:

- [R for Data Science](https://r4ds.hadley.nz/data-transform.html#groups): This provides a great summary of group_by, summarize, and slice_
- [Data School Video on Dplyr](https://www.dataschool.io/dplyr-tutorial-for-faster-data-manipulation-in-r/)
- [A Gentle Introduction](https://bookdown.org/jgscott/DSGI/summaries.html) This just includes summaries, not grouping. 
