---
title: "Summary Tabples"
output:
  html_document:
    toc: true 
    toc_float: true
    toc_collapsed: true
    toc_depth: 3  
layout: default
---

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

## Creating summary tables

We have begun in class summarizing different variables in data frames. But often we are interested in summaries based on different groups. For example, in our climate data analysis, we looked to see what the average precipitation value was in Raleigh, but maybe we want to see the average for every year. We can do this utilizing both the `group_by()` and `summarize()` functions. 

- `group_by()` puts the data frame in "groups" based on a column or columns. This is down "behind the scenes" and doesn't change the data output. But, when your run other functions, like `mutate()` or `summarize()` the groups are honored. 

- `summarize()` allows you to make new columns based on a summary functions -- like the mean, median, or count. `summarize()` uses a similar format to `mutate`, new_column = function. 

For example, if I wanted to get the average (`mean()`) of precipitation  (`precip`) for the different years, you could do the following:

```{r, message = F}
climate <- read.csv('data/raleigh_prism_climate.csv')
library(dplyr)

climate |>
  group_by(year) |>
  summarize(average_precip = mean(precip)) 
```

This first code `group_by` groups the data by the year, so that when you do the `summarize` the average will be taken for each year group. 

You can also make multiple summaries at once by adding into the `summarize()` function. Here, we can add in the standard deviation (`sd()`) and average and standard deviation for the mean temperature Let's also add in a filter for only years in the 90's.

```{r}
climate |>
  filter(year %in% 1990:1999) |>
  group_by(year) |>
  summarize(average_temp = mean(tmean), sd_temp = sd(tmean), average_precip = mean(precip), sd_precip = sd(precip))
```

[**Exercise**]ggplot(wake, aes(x=Year, y = Carbon_monoxide_tons)) +
  geom_line(color = 'cornflowerblue') +
  theme_bw()(#summarizing-data): Which year had the highest *total* precipitation in Raleigh?

### Another summary table example

## Water quality data

In this course, we will exploring historic water quality data from the City of Raleigh. The data is collected from 2008 - 2023 and includes 19 different water quality variables, including water temperature, *E. coli* bacteria, and dissolved oxygen for 18 different stream sites. 

Let's now go back to our City of Raleigh water quality data and use plots to explore the data a bit before we do some summaries. 

First, we need to read in the data into R. This data is stored on our class website, not in your workspace. You can run this code to "read" the data into R. 

```{r}
wq_clean <- read.csv('https://maryglover.github.io/bio331/water_quality_manip/raleigh_wq_clean.csv')
```

Look at the columns to see the different variables that were measured in the streams. 

```{r}
colnames(wq_clean)
```

**Activity: **First, pick one of the different parameters and make a histogram. Record the code in your script and make a note about what you see. 

Now, let's use the data to explore the relationship of temperature and dissolved oxygen (in percent saturation). What do you predict?

Make a scatterplot with Temperature_C on the x asis and do_percent_sat on the y axis. 

```{r}
library(ggplot2)
ggplot(wq_clean, aes(x = Temperature_C, do_percent_sat)) +
  geom_point()
```  

Now, let's see what we notice about the different sites. We can start with exploring the amount of phosphorus at each site using a boxplot.

```{r}
ggplot(data = wq_clean, aes(x = Site, y = Phosphorus_total_mg_L )) +
  geom_boxplot()
```

Now, look at the amount of E. coli in MPN units. What do you notice?

```{r}
ggplot(data = wq_clean, aes(x = Site, y = E_coli_MPN_100mL )) +
  geom_boxplot()
```

Here, you can see that there is an outlier! Use `dplyr` functions to sort the *E. coli* data to look more closely. 

This is an example of a time where good scientific judgement is needed to figure out what to do with this data point. It is *not* appropriate to remove a data point just because it is very different from another. However, there are times when an outlier is due to an error in data collection or recording. It is important to understand the parameter to determine what is going on. This is more complicated when we did not collect the data. In this instance, I was able to talk to the data collector and they indicated that this value for *E. coli * is not a reasonable value and is likely due to a lab issue. In this case, we can remove this data point. 

Run the following code to do that. 

```{r}
wq_clean <- wq_clean |>
 mutate(E_coli_MPN_100mL = gsub('155310', NA, E_coli_MPN_100mL)) |>
  mutate(E_coli_MPN_100mL = as.numeric(E_coli_MPN_100mL))
```

Now, let's look at the boxplot again. 

```{r}
ggplot(data = wq_clean, aes(x = Site, y = E_coli_MPN_100mL)) +
  geom_boxplot()
```

Because we have made some changes in the water quality data, it is a good idea to save the data set. We can do this with the `write.csv()` function. We *should not* rewrite the original data file, but should rename it. Here, we provide the arguments for the data frame we want to save, and the name of the file. 

```{r, eval=FALSE}
write.csv(wq_clean, 'data/raleigh_water_analysis.csv', row_names = FALSE)
```

## Summarizing the water quality data

Now, that we have learned a little about the water quality data, let's used what we learned about grouping and summarizing to learn a little more. 

Let's look first at the Turbidity_NTU data to see how the different streams. First, let's just see what the min and max values are for all the streams.

```{r, eval=FALSE}
max(wq_clean$Turbidity_NTU)
```

What do you notice? Does that work?

In this data, and in many other data sets, there are NA values, meaning not applicable or not available. This is used when we do not have information about a value. We can't put zero when we don't know, because we cannot say that the value is zero. Because there are NAs in our data set, we can't say for sure that the maximum turbidity in the streams is not in a stream that wasn't measured. So the answer for the maximum value is NA. To ignore the NAs to get the maximum turbidity based on the values that we measured, we must add `na.rm = TRUE` to the function. 

```{r, eval=FALSE}
max(wq_clean$Turbidity_NTU, na.rm=TRUE)
```

Now, let's look at the turbidity average for each of the different streams, using the column `Site`.

```{r}
wq_clean |>
  group_by(Site) |>
  summarize(mean_turb = mean(Turbidity_NTU, na.rm = T))
```

What do you notice about the data?

You can use this table to match the Site codes with the stream name. 

```{r, echo = F}
stream_codes <- read.csv('../../data/raleigh_wq_stream_codes.csv', header = F)
knitr::kable(stream_codes)
```

## Homework exercises

- climate data. 
- make a table of the average precip for each month
- average temperature. 

- load in the data for the stream invertebrates. 
- which group collected the most insects total.
- which group collected the most distinct insects
- What insect was the most abundant. 

- do a few more summaries from the water quality data and ask some questions. 
- What do you notice
- what would you like to do but dont know how
- what else might be interesing to look at with the water quality data. 
Ask questions and create summarizes using functions group_by, summarize, etc.
As you are writing code and exploring the data, make sure you are using comments in your script (using “#”) to describe what each step of your code is doing.
Write up a paragraph describing what you learned from the water quality data. You can do this also in the script.

## List of new functions

- `group_by`
- `summarize`
- `sd`
- `na.rm = T`