---
title: "Advanced R tools"
output:
  html_document:
    toc: true 
    toc_float: true
    toc_collapsed: true
    toc_depth: 3  
layout: default
---

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

## Joins

Sometimes it is useful to work with multiple datasets to evaluate a question. It can be helpful to **join** the datasets into one. 

You can do this using dplyr "join" functions. There are a few different join functions based on what rows to include from the two data sets. For example, you can keep every row from both data sets (full_join), only rows that are in both (inner_join), or all rows from one (left_join or right_join) 

![](./07-dplyr_joins.svg)

When doing a join, you must give at least one column to know how to match the two data sets. 

For example, let's go back the macroinvertebrate data from the class from 2024. 


First, we will read in the data from our class website. I will call it `invert`. 

```{r}
invert <- read.csv('https://maryglover.github.io/bio331/open_data/macroinvertebrate_sampling.csv')
```

Here, again we have data for what invertebrates were collected by the class. 
```{r}
head(invert)
```

Macroinvertebrate sampling is often used in streams to measure water quality by looking at how tolerant the taxa are to pollution. If there are many taxa that are intolerant, or cannot live in streams with lots of pollution, that indicates that the stream is relatively healthy. 

Here, we will join in a data file that has the pollution class for different aquatic macroinvertebrates. 


For example, 


The water quality data set includes many water quality metrics. However, you may want to add in other data to use with the water quality data in your analysis. 

For example, one question the City of Raleigh has is how the precipitation might affect the values collected. To answer this question, you have to "join" two data sets. 



let's add a column where we add the full stream name. We will match with the stream code. 

In the exercise, you used separate to make a column for latitude and longitude in the stream. Now let's separate it further with the code and stream name. To make it easier, we will give the code the same column name as in the water quality dataset (Site)


We want to keep all the rows in the water quality data set. So we will do a left join (with the water quality data as the "left" argument)

```{r, eval = F}
library(dplyr)
left_join(water, streams_sep) |>
  head()
```
By default, the function found the column that matched both data frames. You could also specify the argument `by = "Site"` to do the same. 

Now let's say that you only wanted to include the stream name and not the x and y coordinates. You can incorporate pipes with the joins as well. Here is  also give an example of what to do when the join column names do not match. 

```{r, eval = F}
streams |>
  separate(V1, into = c('Code', 'Stream'), sep = '—') |>
  select(Code, Stream) |>
  right_join(water, by = c('Code' = 'Site')) |>
  head()
```

## Dealing with Dates

Dates can be a complicated data format to work with in R. They are often characterized as "characters", however they do have a quantitative nature and can thus be considered as continuous. 

With the water quality data, the date is in the format yyyy-mm-dd. 

The package `lubridate` does this easily. Lubridate has a variety of functions based on the order of the date objects in the string, for example dmy(), myd(), ymd(). 

Our date is in the year, month, date format so we can use the function ymd() to turn the Date column into a date class. 


- raleigh climate make one to go through the entire dataset 

```{r, message=F, eval = F}
library(lubridate)
water_edit<- water_edit |>
  mutate(Date = ymd(Date)) 

head(water_edit)
```

You can see the data itself didn't change, but under Date, the type of data has changed. 

You can also use other lubridate functions to extract out the date components like month(), year(), wday(). For example, you could use year() with mutate() to get a column for the year.

```{r, eval = F}
water_edit |>
  mutate(year = year(Date))|>
  head()
```

[**Exercise**](#plot-with-date) This date format comes in handy when plotting. Choose 1 variable in 1 stream and plot with Date on the y axis


## Exercises
- total number of babies born over time and plot by year
- cycling crashes - put in a date column 
- plot number of crashes each day over time. color the points by the weather
- use the nc_monitoring or the water quality data to answer questiosn

precip <- read.csv('data/precip_stream_sites')

## List of new functions
- lubridate
- join 

## Resources

- [R for Data Science](https://r4ds.hadley.nz/transform)
- more about [dplyr joins](https://www.tidyverse.org/blog/2023/01/dplyr-1-1-0-joins/)
- [Lubridate cheat sheet](https://rawgit.com/rstudio/cheatsheets/main/lubridate.pdf)

