---
title: "Additional practice questions"
output:
  html_document:
    toc: true 
    toc_float: true
    toc_collapsed: true
    toc_depth: 3  
layout: default
---

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

## Set up 

In this practice, we will work with a dataset that comes with `ggplot2` package. 

To load this data, run the following code. 

```{r}
library(ggplot2)
data(mpg)
```

You can now see the top few lines using the `head()` function

```{r}
head(mpg)
```

This dataset include fuel economy data for 38 different car models in 1999 and 2008. We will work with this dataset in the following this practice set. 

## Evaluate the data frame.

In this activity, I will provide hints throughout. Only use the hints if you need them! Try to see what you can do without them first. 

1. What are the column names in the dataset? 
<details>
  <summary>Hint</summary>
    Use the function `colnames()`
</details>
1. How many rows are in the dataset? 
<details>
  <summary>Hint</summary>
    Use the function `nrow()`
</details>
1. What is the average highway mpg for the entire dataset?
<details>
  <summary>Hint</summary>
   You will have to select just the highway `hwy` column using the $ sign. The function for average is `mean()`
</details>
2. What is the minimum city mpg for the entire dataset?
<details>
  <summary>Hint</summary>
   You will have to select just the city mpg `cty` column using the $ sign. The function for average is `min()`</details>

## `dplyr` functions

1. How many different car classes are there?
2. select just the manufacturer, model, year, and hwy mpg.
1. How many are suvs?
1. how many hondas are there in 2008
1. how many hondas or toyotas were there in 1999
1. what car has the lowest highway mpg
1. what car has the highest city mpg. 
1. what car in 1990 had the highest highway mpg. 
1. mutate to get difference in hwy and city mpg

## `ggplot` graphing

1. make a histogram of city mpg
1. make a histogram of hwy mpg
1. make a boxplot of cty mpg and class
1. make a boxplot of hwy mpg and manufacturer
1. geom point of hwy mpg and class, color by manufacturer

```{r}
ggplot(data = mpg, aes(x = class, y = hwy, color = manufacturer, shape = manufacturer))+
  geom_point()
```

1. pick one of the plots and make it pretty
1. change the color
1. add a them
1. chnage the y axis label
1. Add a title

<details>
  <summary>Hint</summary>
  
Hint: put _ on x axis and __ on the y axis

</details>