---
title: "Additional practice questions"
output:
  html_document:
    toc: true 
    toc_float: true
    toc_collapsed: true
    toc_depth: 3  
layout: default
---

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

## Set up 

In this practice, we will work with a dataset that comes with `ggplot2` package. 

To load this data, run the following code. 

```{r}
library(ggplot2)
data(mpg)
```

You can now see the top few lines using the `head()` function

```{r}
head(mpg)
```

This dataset include fuel economy data for 38 different car models in 1999 and 2008. We will work with this dataset in the following this practice set. 

## Evaluate the data frame.

In this activity, I will provide hints throughout. Only use the hints if you need them! Try to see what you can do without them first. 

1. What are the column names in the dataset? 
<details>
  <summary>Hint</summary>
    Use the function `colnames()`
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    colnames(mpg)
    ```
</details>
1. How many rows are in the dataset? 
<details>
  <summary>Hint</summary>
    Use the function `nrow()`
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    nrow(mpg)
    ```
</details>
1. What is the average highway mpg for the entire dataset?
<details>
  <summary>Hint</summary>
   You will have to look at just the highway `hwy` column using the $ sign. The function for average is `mean()`
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    mean(mpg$hwy)
    ```
</details>
2. What is the minimum city mpg for the entire dataset?
<details>
  <summary>Hint</summary>
   You will have to select just the city mpg `cty` column using the $ sign. The function for average is `min()`</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    min(mpg$cty)
    ```
</details>

## `dplyr` functions

1. What are the different car classes that are in the dataset?
<details>
  <summary>Hint1</summary>
   You will need to load in the `dplyr` package before you can use any of the functions. `library(dplyr)`
   </details>
<details>
  <summary>Hint2</summary>
    You can use the `distinct` function on the column `class`.
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    mpg |>
      distinct(class)
    ```
</details>
2. Select the columns for just the manufacturer, model, year, and highway mpg.
<details>
  <summary>Hint</summary>
   Use the `select()` function and type the column names exactly as they appear in `colnames()`
   </details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    mpg |>
      select(manufacturer, model, year, hwy)
    ```
</details>
1. How many of the cars in the dataset are suvs?
<details>
  <summary>Hint</summary>
 You need use the `filter()` function to show only suvs from the `class` column. You can also add the `nrow()` function to get the number of rows. 
 </details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    mpg |>
      filter(class == 'suv')
    
    mpg |>
      filter(class == 'suv')|>
      nrow()
    ```
</details>
1. How many Hondas were there in 2008?
<details>
  <summary>Hint</summary>
  You will have to filter for *both* honda and 2008. 
  </details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    mpg|>
      filter(manufacturer== 'honda', year == '2008')
    ```
</details>
1. How many Hondas *or* Toyotas were there in 1999?
<details>
  <summary>Hint</summary>
   You will have to filter for honda or toyota using the "|" symbol between your filtering for manufacturer **and** filter the year for 1999. 
   </details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    mpg |>
      filter(manufacturer == 'honda' | manufacturer == 'toyota', year == 1999)
    ```
</details>
1. What car has the lowest highway mpg?
<details>
  <summary>Hint</summary>
   Use the `arrange()` function on the `hwy` column.
   </details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    mpg |>
      arrange(hwy)
    ```
</details>
1. What car has the highest city mpg?
<details>
  <summary>Hint</summary>
   You will `arrange()` based on the cty column. To go from greatest to least you can put a - sign in front of the column. 
   
   You can also add `head()` to just show the first few rows 
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
mpg |>
      arrange(-cty)
    
    mpg |>
      arrange(-cty) |>
      head()
    ```
</details>
1. What car had the highest highway mpg in 1999? 
<details>
  <summary>Hint</summary>
   You will `arrange()` based on the hwy column. To go from greatest to least you can put a - sign in front of the column. You also need to `filter()` for 1999. 
</detais>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
  mpg |>
      arrange(-hwy) |>
      filter(year == 1999)
    ```
</details>
1. Mutate the table to make a new column with the difference of highway mpg and city mpg. 
<details>
  <summary>Hint</summary>
  You will use the `mutate()` function which expects a new column name = the calculation from the columns in the data frame. i.e. mutate(NEWCOL = COL1 + COL2). 
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    mpg |>
      mutate(diff = hwy - cty)
    
    ```
</details>

## `ggplot` graphing

1. Make a histogram of city mpg. 
<details>
  <summary>Hint</summary>
  Set up the plot with the following format `ggplot(data, aes (x = )`. Then add the `geom_histogram()` layer.
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    ggplot(mpg, aes(x = cty)) +
      geom_histogram()
    ```
</details>
1. Make a histogram of highway mpg.
<details>
  <summary>Hint</summary>
Change your previous code to have the highway `hwy` column instead of `cty`
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    ggplot(mpg, aes(x = hwy)) +
      geom_histogram()
    ```
</details>
1. Make a boxplot of city mpg (y axis) for each class (x axis)
<details>
  <summary>Hint</summary>
  Set up the plot with the following format `ggplot(data, aes (x = , y =  )`. Then add the `geom_boxplot()` layer.
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    ggplot(mpg, aes(x = class, y = cty)) +
      geom_boxplot()
    ```
</details>
1. Make a boxplot of highway mpg for each manufacturer. 
<details>
  <summary>Hint</summary>
  Change the y axis to `hwy` instead of `cty`. 
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    ggplot(mpg, aes(x = class, y = hwy)) +
      geom_boxplot()
    ```
</details>
1. Make a scatterplot for the highway mpg (y axis) and class (x axis). Change the color of the points so that each manufacturer is a different color. <details>
<details>
  <summary>Hint2</summary>
To change the color, you must specify color = *inside* and aes() function in geom_point. The color should = the *column* you want to color by. 
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    ggplot(mpg, aes(x = class, y = hwy)) +
      geom_point(aes(color = manufacturer))
    ```
</details>
1. Make a scatterplot of the relationship between the size of the engine (the `displ` column) on the x axis and the city gas mileage on the y axis. 
<details>
  <summary>Hint</summary>
  Set up the plot in the same was as the last plot, changing the columns for the x and y axis to `displ` and `cty`, respectively
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    ggplot(mpg, aes(x = displ, y = cty)) +
      geom_point()
    ```
</details>
1. Using the previous scatterplot, change the color of the points to whatever color you prefer. 
<details>
  <summary>Hint</summary>
  Using the last plot, add a `color =` into the `geom_point` function. Change it to any color you would like. </details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    ggplot(mpg, aes(x = displ, y = cty)) +
      geom_point(color = 'slateblue')
    ```
</details>
1. Using the previous scatterplot, add a theme. 
<details>
  <summary>Hint</summary>
  Using the last plot, add any theme_* that you prefer. For example, `theme_minimal()`. </details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    ggplot(mpg, aes(x = displ, y = cty)) +
      geom_point(color = 'slateblue') +
      theme_bw()
    ```
</details>
1. Using the previous scatterplot, change the x and y axis labels
<details>
  <summary>Hint</summary>
  Using the last plot, use the function `labs()` to add an x and y label. Label names should be in quotations. </details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    ggplot(mpg, aes(x = displ, y = cty)) +
      geom_point(color = 'slateblue')+
      theme_bw()+
      labs(x = 'Engine Size', y = 'City Gas Mileage')
    ```
</details>
1. Using the previous scatterplot, add a title. 
<details>
  <summary>Hint</summary>
  Using the last plot, use the function `labs()` to add a title. </details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    ggplot(mpg, aes(x = displ, y = cty)) +
      geom_point(color = 'slateblue')+
      theme_bw()+
      labs(x = 'Engine Size', y = 'City Gas Mileage', title = "Gas mileage decreases with engine size")
    ```
</details>