---
title: "Additional practice questions exam 2"
output:
  html_document:
    toc: true 
    toc_float: true
    toc_collapsed: true
    toc_depth: 3  
layout: default
---

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

## Set up 

In this practice, we will work with a dataset that comes with `ggplot2` package. 

To load this data, run the following code. 

```{r}
library(ggplot2)
data(mpg)
```

You can now see the top few lines using the `head()` function

```{r}
head(mpg)
```

This dataset include fuel economy data for 38 different car models in 1999 and 2008. We will work with this dataset in the following this practice set. 

In this activity, I will provide hints throughout. Only use the hints if you need them! Try to see what you can do without them first. 

## Summary tables

1. Before you begin the summaries, make sure you load the package to do summaries!
<details>
  <summary>Hint</summary>
   Load in the `dplyr` package
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    library(dplyr)
    ```
</details>
1. Make a summary table to include show the *average* city gas mileage, `cty`, for each vehicle class, `class`.
<details>
  <summary>Hint</summary>
    Group by class and summarise based on cty
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    mpg |>
      group_by(class)|>
      summarize(city_mean = mean(cty))
    ```
</details>
1. Make a summary table to show the *maximum* highway gas mileage (`hwy`) for each manufacturer (`manufacturer`).
<details>
  <summary>Hint</summary>
    Group by the manufacturer and summarize to calculate the `max()` for the highway mileage. 
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    mpg |> 
      group_by(manufacturer) |>
      summarize(hwy_max = max(hwy))
    ```
</details>
1. Make a summary table to show the city gas mileage *and* the highway gas mileage for each year.
<details>
  <summary>Hint</summary>
   Group by the year. Then add two different summaries in `summarize()`
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
   mpg|>
      group_by(year) |>
      summarize(city_mean = mean(cty), hwy_mean = mean(hwy))
    ```
</details>
1. Make a summary table to show the mean city gas mileage for each class type *and* year.
<details>
  <summary>Hint</summary>
   Group by the year *and* class
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
   mpg|>
      group_by(class, year) |>
      summarize(city_mean = mean(cty))
    ```
1. Filter the summary table you just make to only include the year 2008. 
<details>
  <summary>Hint</summary>
  Add a filter to the code you just used to make the table after a pipe. 
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
   mpg|>
      group_by(class, year) |>
      summarize(city_mean = mean(cty)) |>
      filter(year == 2008)
    ```
1. Make a summary table to display the median city mileage for every manufacturer. Then, sort to find which manufacturer has the highest gas mileage. 
<details>
  <summary>Hint</summary>
    Make the summary table and then `arrange()` by the column you made in the `summarize` function
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    mpg |>
      group_by(manufacturer) |>
      summarize(city_median = median(cty))|>
      arrange(-city_median)
    ```
</details>

## Statistics

1. Do a statistical test to determine if the year the car was manufactured affected the city gas mileage. 
<details>
  <summary>Hint</summary>
   Use the function `lm()` with `cty` as the response and `year` as the predictor
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    year_lm <- lm(data = mpg, cty ~ year )
    ```
</details>
2. View the statistical test. Is there a significant difference in the gas mileage for the different years?
<details>
  <summary>Hint</summary>
  Use the function `summary()` to view the statistical model
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
  summary(year_lm)
    ```
    There is no difference between the different years. 
  </details>
  
1. Do a statistical test to see if the highway gas mileage is predicted by the vehicle class 
<details>
  <summary>Hint</summary>
   Use the function `lm()` with `hwy` as the response and `class` as the predictor
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
    class_lm <- lm(data = mpg, hwy ~ class )
    ```
</details>
2. View the statistical test. Is there a significant difference in the gas mileage based on vehicle class?
<details>
  <summary>Hint</summary>
  Use the function `summary()` to view the statistical model
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
  summary(class_lm)
    ```
    There is a significant effect of vehicle class on highway gas mileage.
  </details>
1. Make a boxplot to show the highway gas mileage for each vehicle class. 
<details>
  <summary>Hint</summary>
  Use your ggplot functions to make a boxplot with the mpg data, with `class` on x axis and `hwy` on y axis. 
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
  ggplot(mpg, aes(x=class, y = hwy)) +
      geom_boxplot() +
      theme_classic()
    ```
    There is a significant effect of vehicle class on highway gas mileage.
  </details>
1. Based on the plot, what class had the lowest highway mileage. 
1. Perform a statistical test to see if city gas mileage predicts highway gas mileage. 
  <details>
  <summary>Hint</summary>
Use the `lm()` function with cty as the predictor and hwy as the response
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
mpg_lm <- lm(data = mpg, hwy ~ cty)
    ```
  </details>
1. Is the result significant?
<details>
  <summary>Hint</summary>
Use the function `summary` to view the results. 
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
summary(mpg_lm)
    ```
    There is a significant effect of city mileage on highway milegae. 
  </details>
1. Plot a scatterplot of the city and highway gas mileage and add the best fit line. 
<details>
  <summary>Hint</summary>
  Use your ggplot functions to make a boxplot with the mpg data, with `cty` on x axis and `hwy` on y axis. Add a `geom_smooth()` layer
</details>
<details>
  <summary>Answer</summary>
    ```{r, eval = F}
  ggplot(mpg, aes(x=cty, y = hwy)) +
      geom_point() +
      geom_smooth(method = 'lm') +
      theme_classic()
    ```
  </details>

## Dates

